---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## source
https://timogrossenbacher.ch/2016/12/beautiful-thematic-maps-with-ggplot2-only/

## clearing workspace
Clear workspace and install necessary packages
This is just my usual routine: Detach all packages, remove all variables in the global environment, etc, and then load the packages. Saves me a lot of headaches.

```{r clear}
knitr::opts_chunk$set(
    out.width = "100%",
    dpi = 300,
    fig.width = 8,
    fig.height = 6,
    fig.path = 'https://timogrossenbacher.ch/wp-content/uploads/2016/12/tm-',
    strip.white = T,
    dev = "png",
    dev.args = list(png = list(bg = "transparent"))
)

remove(list = ls(all.names = TRUE))

detachAllPackages <- function() {
  basic.packages.blank <-  c("stats", 
                             "graphics", 
                             "grDevices", 
                             "utils", 
                             "datasets", 
                             "methods", 
                             "base")
  basic.packages <- paste("package:", basic.packages.blank, sep = "")

  package.list <- search()[ifelse(unlist(gregexpr("package:", search())) == 1, 
                                  TRUE, 
                                  FALSE)]

  package.list <- setdiff(package.list, basic.packages)

  if (length(package.list) > 0)  for (package in package.list) {
    detach(package, character.only = TRUE)
    print(paste("package ", package, " detached", sep = ""))
  }
}

detachAllPackages()


if (!require(rgeos)) {
  install.packages("rgeos", repos = "http://cran.us.r-project.org")
  require(rgeos)
}
if (!require(rgdal)) {
  install.packages("rgdal", repos = "http://cran.us.r-project.org")
  require(rgdal)
}
if (!require(raster)) {
  install.packages("raster", repos = "http://cran.us.r-project.org")
  require(raster)
}
if(!require(ggplot2)) {
  install.packages("ggplot2", repos="http://cloud.r-project.org")
  require(ggplot2)
}
if(!require(viridis)) {
  install.packages("viridis", repos="http://cloud.r-project.org")
  require(viridis)
}
if(!require(dplyr)) {
  install.packages("dplyr", repos = "https://cloud.r-project.org/")
  require(dplyr)
}
if(!require(gtable)) {
  install.packages("gtable", repos = "https://cloud.r-project.org/")
  require(gtable)
}
if(!require(grid)) {
  install.packages("grid", repos = "https://cloud.r-project.org/")
  require(grid)
}
if(!require(readxl)) {
  install.packages("readxl", repos = "https://cloud.r-project.org/")
  require(readxl)
}
if(!require(magrittr)) {
  install.packages("magrittr", repos = "https://cloud.r-project.org/")
  require(magrittr)
}
```

## General ggplot2 theme for map

First of all, I define a generic theme that will be used as the basis for all of the following steps. It's based on theme_minimal and basically resets all the axes. It also defined a very subtle grid and a warmgrey background, which gives it some sort of paper map feeling, I find.

The font used here is Ubuntu Regular – adapt to your liking, but the font must be installed on your OS.

```{r general theme, echo=FALSE}
theme_map <- function(...) {
  theme_minimal() +
  theme(
    text = element_text(family = "Ubuntu Regular", color = "#22211d"),
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    # panel.grid.minor = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.major = element_line(color = "#ebebe5", size = 0.2),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#f5f5f2", color = NA), 
    panel.background = element_rect(fill = "#f5f5f2", color = NA), 
    legend.background = element_rect(fill = "#f5f5f2", color = NA),
    panel.border = element_blank(),
    ...
  )
}
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Data sources

For this choropleth, I used three data sources:

Thematic data: Average age per municipality as of end of 2015. The data is freely available from The Swiss Federal Statistical Office (FSO) and included in the input folder.

Municipality geometries: The geometries do not show the political borders of Swiss municipalities, but the so-called "productive" area, i.e., larger lakes and other "unproductive" areas such as mountains are excluded. This has two advantages: 1) The relatively sparsely populated but very large municipalities in the Alps don't have too much visual weight and 2) it allows us to use the beautiful raster relief of the Alps as a background. The data are also from the FSO, but not freely available. You could also use the freely available political boundaries of course. I was allowed to republish the Shapefile for this educational purpose (also included in the input folder). Please stick to that policy.

Relief: This is a freely available GeoTIFF from The Swiss Federal Office of Topography (swisstopo).

## Read in data and preprocess
## Read in geodata

Here, the geodata is loaded using rgeos / rgdal standard procedures. It is then "fortified", i.e. transformed into a ggplot2-compatible data frame (the fortify-function is part of ggplot2). Also, the thematic data is joined using the bfs_id field (each municipality has a unique one).

```{r data}
data <- read.csv("input/avg_age_15.csv", stringsAsFactors = F)
```
```{r reading OGR}
gde_15 <- readOGR("input/geodata/gde-1-1-15.shp", layer = "gde-1-1-15")
## OGR data source with driver: ESRI Shapefile 
## Source: "input/geodata/gde-1-1-15.shp", layer: "gde-1-1-15"
## with 2324 features
## It has 2 fields
# set crs to ch1903/lv03, just to make sure  (EPSG:21781)
crs(gde_15) <- "+proj=somerc +lat_0=46.95240555555556 
+lon_0=7.439583333333333 +k_0=1 +x_0=600000 +y_0=200000 
+ellps=bessel +towgs84=674.374,15.056,405.346,0,0,0,0 +units=m +no_defs"
# fortify, i.e., make ggplot2-compatible
map_data_fortified <- fortify(gde_15, region = "BFS_ID") %>% 
  mutate(id = as.numeric(id))
# now we join the thematic data
map_data <- map_data_fortified %>% left_join(data, by = c("id" = "bfs_id"))

# whole municipalities
gde_15_political <- readOGR("input/geodata/g1g15.shp", layer = "g1g15")
## OGR data source with driver: ESRI Shapefile 
## Source: "input/geodata/g1g15.shp", layer: "g1g15"
## with 2328 features
## It has 20 fields
crs(gde_15_political) <- "+proj=somerc +lat_0=46.95240555555556 
+lon_0=7.439583333333333 +k_0=1 +x_0=600000 +y_0=200000 
+ellps=bessel +towgs84=674.374,15.056,405.346,0,0,0,0 +units=m +no_defs"
map_data_political_fortified <- fortify(gde_15_political, region = "GMDNR") %>% 
  mutate(id = as.numeric(id))
map_data_political <- map_data_political_fortified %>% left_join(data, by = c("id" = "bfs_id"))
map_data_political <- map_data_political[complete.cases(map_data_political),]
# read in background relief
relief <- raster("input/geodata/02-relief-georef-clipped-resampled.tif")
relief_spdf <- as(relief, "SpatialPixelsDataFrame")
# relief is converted to a very simple data frame, 
# just as the fortified municipalities.
# for that we need to convert it to a 
# SpatialPixelsDataFrame first, and then extract its contents 
# using as.data.frame
relief <- as.data.frame(relief_spdf) %>% 
  rename(value = `X02.relief.georef.clipped.resampled`)
# remove unnecessary variables
rm(relief_spdf)
rm(gde_15)
rm(map_data_fortified)
rm(map_data_political_fortified)

```

## Basic map
What follows now is a very basic map with the municipalities rendered with geom_polygon and their outline with geom_path. I don't even define a color scale here, it just uses ggplot2's default continuous color scale, because avg_age_15 is a continuous variable.

Because the geodata are in a projected format, it is important to use coord_equal() here, if not, Switzerland would be distorted.

```{r basic plot}
p <- ggplot() +
    # municipality polygons
    geom_polygon(data = map_data, aes(fill = avg_age_15, 
                                      x = long, 
                                      y = lat, 
                                      group = group)) +
    # municipality outline
    geom_path(data = map_data, aes(x = long, 
                                   y = lat, 
                                   group = group), 
              color = "white", size = 0.1) +
    coord_equal() +
    # add the previously defined basic theme
    theme_map() +
    labs(x = NULL, 
         y = NULL, 
         title = "Switzerland's regional demographics", 
         subtitle = "Average age in Swiss municipalities, 2015", 
         caption = "Geometries: ThemaKart, BFS; Data: BFS, 2016")
p
```

## Better color scale

See how I reuse the previously defined p-object and just add the continuous viridis scale from the same named package. All of a sudden the map looks more aesthetic and regional patterns are already visible in this linear scale. For example one can see that the municipalities in the south and in the Alps (where there are a lot of gaps, the unproductive areas I talked about) seem to have an older-than-average population (mainly because young people move to the cities for work etc.).

```{r viridis color scale}
q <- p + scale_fill_viridis(option = "magma", direction = -1)
q
```


## Horizontal legend

Also I think one could save some space by using a horizontal legend at the bottom of the plot.

```{r horizontal legend}
q <- p +
  # this is the main part
  theme(legend.position = "bottom") +
  scale_fill_viridis(
    option = "magma", 
    direction = -1,
    name = "Average age",
    # here we use guide_colourbar because it is still a continuous scale
    guide = guide_colorbar(
      direction = "horizontal",
      barheight = unit(2, units = "mm"),
      barwidth = unit(50, units = "mm"),
      draw.ulim = F,
      title.position = 'top',
      # some shifting around
      title.hjust = 0.5,
      label.hjust = 0.5
  ))
q
```


## Discrete classes with quantile scale

I am still not happy with the color scale because I think regional patterns could be made more clearly visible. For that I break up the continuous avg_age_15 variable into 6 quantiles (remember your statistics class?). The effect of that is that I now have about the same number of municipalities in each class.

```{r discrete quantile scale}

no_classes <- 6
labels <- c()

quantiles <- quantile(map_data$avg_age_15, 
                      probs = seq(0, 1, length.out = no_classes + 1))

# here I define custom labels (the default ones would be ugly)
labels <- c()
for(idx in 1:length(quantiles)){
  labels <- c(labels, paste0(round(quantiles[idx], 2), 
                             " – ", 
                             round(quantiles[idx + 1], 2)))
}
# I need to remove the last label 
# because that would be something like "66.62 - NA"
labels <- labels[1:length(labels)-1]

# here I actually create a new 
# variable on the dataset with the quantiles
map_data$avg_age_15_quantiles <- cut(map_data$avg_age_15, 
                                     breaks = quantiles, 
                                     labels = labels, 
                                     include.lowest = T)

p <- ggplot() +
    # municipality polygons (watch how I 
   # use the new variable for the fill aesthetic)
    geom_polygon(data = map_data, aes(fill = avg_age_15_quantiles, 
                                      x = long, 
                                      y = lat, 
                                      group = group)) +
    # municipality outline
    geom_path(data = map_data, aes(x = long, 
                                   y = lat, 
                                   group = group), 
              color = "white", size = 0.1) +
    coord_equal() +
    theme_map() +
    labs(x = NULL, 
         y = NULL, 
         title = "Switzerland's regional demographics", 
         subtitle = "Average age in Swiss municipalities, 2015", 
         caption = "Geometries: ThemaKart, BFS; Data: BFS, 2016") +
  # now the discrete-option is used, 
  # and we use guide_legend instead of guide_colourbar
  scale_fill_viridis(
    option = "magma",
    name = "Average age",
    discrete = T,
    direction = -1,
    guide = guide_legend(
     keyheight = unit(5, units = "mm"),
     title.position = 'top',
     reverse = T
  ))
p
```


Wow! Now that is some regional variability ;-). But there is still a huge caveat: In my opinion, quantile scales are optimal at showing intra-dataset-variability, but sometimes this variability can be exaggerated. Most of the municipalities here are in the region between 39 and 43 years. The second caveat is that the legend looks somehow ugly with all these decimals, and that people are probably having problems interpreting such differently sized classes. That's why I am trying "pretty breaks" in the next step, and this is basically also what you see in almost all choropleths used for (data-)journalistic purposes.


## Pretty breaks
```{r pretty breaks}
# here I define equally spaced pretty breaks - 
# they will be surrounded by the minimum value at 
# the beginning and the maximum value at the end. 
# One could also use something like c(39,39.5,41,42.5,43), 
# this totally depends on the data and your personal taste.
pretty_breaks <- c(39,40,41,42,43)
# find the extremes
minVal <- min(map_data$avg_age_15, na.rm = T)
maxVal <- max(map_data$avg_age_15, na.rm = T)
# compute labels
labels <- c()
brks <- c(minVal, pretty_breaks, maxVal)
# round the labels (actually, only the extremes)
for(idx in 1:length(brks)){
  labels <- c(labels,round(brks[idx + 1], 2))
}

labels <- labels[1:length(labels)-1]
# define a new variable on the data set just as above
map_data$brks <- cut(map_data$avg_age_15, 
                     breaks = brks, 
                     include.lowest = TRUE, 
                     labels = labels)

brks_scale <- levels(map_data$brks)
labels_scale <- rev(brks_scale)

p <- ggplot() +
    # municipality polygons
    geom_polygon(data = map_data, aes(fill = brks, 
                                      x = long, 
                                      y = lat, 
                                      group = group)) +
    # municipality outline
    geom_path(data = map_data, aes(x = long, 
                                   y = lat, 
                                   group = group), 
              color = "white", size = 0.1) +
    coord_equal() +
    theme_map() +
    theme(legend.position = "bottom") +
    labs(x = NULL, 
         y = NULL, 
         title = "Switzerland's regional demographics", 
         subtitle = "Average age in Swiss municipalities, 2015", 
         caption = "Geometries: ThemaKart, BFS; Data: BFS, 2016")
q <- p +
    # now we have to use a manual scale, 
    # because only ever one number should be shown per label
    scale_fill_manual(
          # in manual scales, one has to define colors, well, manually
          # I can directly access them using viridis' magma-function
          values = rev(magma(6)),
          breaks = rev(brks_scale),
          name = "Average age",
          drop = FALSE,
          labels = labels_scale,
          guide = guide_legend(
            direction = "horizontal",
            keyheight = unit(2, units = "mm"),
            keywidth = unit(70 / length(labels), units = "mm"),
            title.position = 'top',
            # I shift the labels around, the should be placed 
            # exactly at the right end of each legend key
            title.hjust = 0.5,
            label.hjust = 1,
            nrow = 1,
            byrow = T,
            # also the guide needs to be reversed
            reverse = T,
            label.position = "bottom"
          )
      )

q
```


Now we have classes with the ranges 33.06 to 39, 39 to 40, 40 to 41, and so on... So four classes are of the same size and the two classes with the extremes are differently sized. One option to communicate this is to make their respective legend keys wider than usual. ggplot2 doesn't have a standard option for that, so I had to dig deep into the underlying grid package and extract the relevant grobs and change their widths. All of the following numbers are the result of trying and trying around. I have not yet fully understood how that system actually works and certainly, it could be made more versatile. Something for next christmas...


